Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-07 _tex/paper2.tex
Built site for gh-pages

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key CJ experimental design features,
such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model to analyze dichotomous
pairwise comparison data. Section~\ref{sec-discussion} discusses the
findings, limitations, and challenges and explores avenues for future
research. Finally, Section~\ref{sec-conclusion} summarizes the study's
conclusions.

@@

Thurstone noted from the outset that his general form, described in
Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
problem}. Specifically, the model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep[pp.~267]{Thurstone_1927b}. For instance, in a CJ experiment with
five texts, the general form would require estimating \(20\) parameters:
five modal discriminal processes, five discriminal dispersions, and
\(10\) correlations --one per comparison (see
Table~\ref{tbl-thurstone_cases}). However, a single judge could only
provide \({5 \choose 2} = 10\) unique comparisons, an insufficient data
set to estimate the required parameters.

To address this issue and facilitate the practical implementation of the
theory, Thurstone developed five cases derived from this general form,
each progressively incorporating additional simplifying assumptions. In
Case I, Thurstone postulated that pairs of stimuli would maintain a
constant correlation across all comparisons. In Case II, he allowed
multiple judges to undertake comparisons instead of confining
evaluations to a single judge. In Case III, he posited that there was no
correlation between stimuli. In Case IV, he assumed that the stimuli
exhibited similar dispersions. Finally, in Case V, he replaced this
assumption with the condition that stimuli had equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes the assumptions
of the general form and the five cases. For a detailed discussion of
these cases and their progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248--253]{Bramley_2008}.

@@

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement, and without providing clear guidance on
how to use Case V's measurement scores for hypothesis testing.
Specifically, he cautioned that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, as the case
imposes the most extensive set of simplifying assumptions
\citep{Bramley_2008, Kelly_et_al_2022} (see
Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
goal was to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b} and ``allocate the compared stimuli on
this continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not address
how to use the resulting measurement scores for hypothesis testing.
Thus, given these limitations, it is surprising that CJ research has
predominantly relied on Case V to measure different traits, raising
significant concerns about the reliability and validity of such
measurements in contexts where the case's assumptions may not hold
\citep{Kelly_et_al_2022, Andrich_1978}. Furthermore, although the CJ
practice has attempted to address the issue of hypothesis testing by
using the scores' point estimates or their transformations, a critical
question remains: Is this approach suitable for addressing these
inquiries?

@@

As previously discussed, Case V remains the most widely used model in CJ
literature. This preference primarily stems from the BTL model, which
provides a simplified statistical representation of the case. The BTL
model mirrors the assumptions of Case V, with one notable distinction:
while Case V assumes a Normal distribution for the stimuli' discriminal
processes, the BTL model uses the more mathematically tractable Logistic
distribution \citep{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
impact on the model's estimation or interpretation because the
discriminal process scale is arbitrary up to a non-monotonic
transformation \citep{vanderLinden_et_al_2017_I, McElreath_2021}.
Furthermore, this limited impact is supported by the fact that the
Normal and Logistic distributions exhibit analogous statistical
properties, differing only by a scaling factor of approximately \(1.7\).

@@

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent texts with different
dispersions or outlying observations, the common CJ practice of
excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
may unintentionally discard valuable information, introducing bias into
the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases remain unpredictable, as they
depend on which stimuli researchers exclude from the analysis.

@@

Building on the previous section, it is clear that, researchers
typically rely on the BTL model to measure a trait and place the
compared stimuli along its continuum \citep{Thurstone_1927b}.
Additionally, the CJ literature shows that researchers frequently use
point estimates of BTL scores or their transformations to conduct
further analyses or hypothesis tests. For example, researchers have used
these scores to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct further analyses or hypothesis tests, as this practice
can undermine the resulting statistical conclusions. A key consideration
is that BTL scores are parameter estimates that inherently carry
uncertainty (measurement error). Ignoring this uncertainty can bias the
analysis and reduce the precision of hypothesis tests. The direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. Furthermore,
the reduced precision in hypothesis tests diminishes their statistical
power, increasing the likelihood of committing type-I or type-II errors
\citep{McElreath_2020}. Figure~\ref{fig-measurement_error} illustrates
these issues, showing that neglecting measurement error
\((\sigma_{T})\)--by relying solely on outcome averages--reduces the
precision of a predictor's estimated effect \(\beta_{X}\), as indicated
by the solid lines.

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/measurement_error.png}

\caption{\label{fig-measurement_error}The effect of outcome uncertainty
\(\sigma_{T}\) on the estimation of an effect \(\beta_{X}\) linked to
the predictor. The example assumes a sample size \(n=100\) and an
uncertainty that increases from left to right.}

@@

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, the SCM incorporates a key design feature of CJ
experiments: the judges' biases \((B_{kj})\). This extension builds on
the arguments presented in Section~\ref{sec-theory-issue1b}, contending
that the discriminal difference becomes an observable outcome only
through judges' perceptions. Given that such perceptions may be
imperfect--and that each judge may carry some degree of bias
\citep[see][]{Pollitt_et_al_2003, vanDaal_et_al_2016}--it is reasonable
that judges' perceptions (bias) should be treated as an integral
component of the CJ system from the outset, as this leads to a more
accurate representation of the data-generating process underlying the
pairwise comparisons. All these variables constitute the preliminary set
of endogenous variables,
\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\), while the
preliminary set of structural equations, \(F = { f_{O}, f_{D} }\),
captures the non-parametric dependencies among them.

@@

Additionally, we construct the matrix \(R\) to map each row of the
\(IA\) matrix with a corresponding row from the \(JK\) matrix. Thus,
this matrix has \(n\) rows and \(6\) columns, where
\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
coefficient, which quantifies the total number of unique comparisons
possible between every pair of stimuli generated by each pair of
individuals in the population. Thus, we define the matrix as follows:

@@

In summary, the SCM~\ref{fig-cj15_scm} and DAG~\ref{fig-cj15_dag} extend
Thurstone's general form to address several limitations of the BTL
model. These extensions account for judge biases (see
Section~\ref{sec-theory-theoretical_P1}), reflect the hierarchical
structure of stimuli (see Section~\ref{sec-theory-theoretical_P3}),
incorporate measurement error in trait estimation and hypothesis testing
(also see Section~\ref{sec-theory-theoretical_P3}), and clarify the role
of the sample and comparison mechanisms in CJ experiments (see
Section~\ref{sec-theory-theoretical_SC}). However, they do not resolve
concerns about the assumption of equal dispersions among stimuli
discussed in Section~\ref{sec-theory-issue1a}. Because this concern
relates to the statistical assumption underlying the distribution of the
discriminal process, we develop a formal statistical model to address it
in the next section.

@@

Note that while the joint distribution in Equation (\ref{eq-mat51})
includes the probability distributions of the sampling and comparison
mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
\(P(Z_{J})\)--we omit these probabilities from the models presented in
Figure~\ref{fig-cj16}. We make this omission because, although these
distributions contribute to the overall joint distribution of the data,
they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
This independence stems from the random selection designs that govern
\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
\(Z_{J}\)\footnote{Randomization ensures that data---and, by extension,
  an estimator---satisfies several key identification properties, such
  as common support, no interference, and consistency. The most critical
  property, however, is the elimination of confounding.
  \emph{Confounding} occurs when an external variable, such as
  \(X_{I}\), simultaneously influences both the outcome (e.g.,
  \(O_{R}\)) and a variable of interest (e.g., \(X_{I}\) or \(S\)),
  resulting in spurious associations between the two
  \citep{Everitt_et_al_2010}. Randomization ensure the absence of
  confounding by effectively decoupling the association between the
  variable of interest and any other variable, except for the outcome
  itself. For a more detailed discussion on the benefits of
  randomization, see \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
  \citet{Neal_2020}, and \citet{Hernan_et_al_2020}.}.

Second, we define \(D_{R}\) as the difference between the discriminal
processes of two texts--\(T_{IA}[i, a]\) and
\(T_{IA}[h, b]\)--representing the underlying written-quality trait of
the texts, plus the corresponding repeated judge biases,
\(B_{JK}[j, k]\). Note that if we assume that \(B_{JK}[j,k]\) reflects
the difference in stimulus-specific biases, i.e.,
\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can re-write the
discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation} This formulation reveals that the discriminal difference
captures a \emph{pure interaction effect}, in which neither the texts'
discriminal processes nor the judges' biases alone determine the
outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
terms, the discriminal processes of the stimuli become an observable
outcome only through the judges' perceptions (biases). To clarify, the
square brackets in \(D_{R}\) simply indicate the relevant indices for
each trait vector and do not imply any subsetting of the data.

@@

Fourth, we define \(T_{I}\) as a linear combination of the effects of
relevant student-related variables, \(\beta_{XI} X_{I}\), and students'
idiosyncratic errors, \(e_{I}\). As before, we assume that each element
of \(e_{I}\) is \(iid\) Normal with a mean of zero. However, we consider
the standard deviation, \(s_{XI}\), varies depending on the teaching
method group to which each student belongs. Given the example in
Section~\ref{sec-theoretical}, where the teaching method
\(X_{I} = \{1,2\}\), the model sets the following constraint to anchor
the scale of the latent trait \citep{Depaoli_2021}:
\(\sum_{g=1}^{2} s_{XI}[g]/2 = 1\). Note that this parametrization of
\(e_{I}\) relaxes the equal dispersion assumption for the stimuli,
effectively addressing the concerns raised in
Section~\ref{sec-theory-issue1a}.

@@

Finally, we use \emph{Bayesian inference methods} to convert
model~\ref{fig-cj16_stat} into a practical statistical tool for
analyzing paired comparison data. Bayesian inference offers three key
advantages for this purpose. First, it handles complex and
overparameterized models, where the number of parameters exceeds the
number of observations \citep{Baker_1998, Kim_et_al_1999}--such as the
one we implement. Second, it incorporates prior information to constrain
parameter estimates within plausible bounds, thereby mitigating
estimation issues like non-convergence or improper solutions that often
affect frequentist methods \citep{Martin_et_al_1975, Seaman_et_al_2011}.
In our implementation, we use priors to define the error distribution,
which sets the scale of the latent variables \citep{Depaoli_2014}.
Third, Bayesian inference supports robust inferences from small samples,
where the asymptotic properties underlying frequentist methods are less
reliable \citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}.
This feature is particularly relevant in CJ experiments, where
researchers often collect large volumes of paired comparisons but work
with relatively small samples of judges, stimuli, and individuals to
test hypotheses.

@@

This study extends Thurstone's general form through a systematic,
integrated approach that combines Causal and Bayesian inference methods.
It begins by formulating the extension as a conceptual model,
represented by an SCM and a DAG. This formulation integrates Thurstone's
core theoretical principles, including the distribution of the
discriminal process, with key CJ experimental design features, such as
the sampling and comparison mechanisms. The study then translates the
SCM into a bespoke statistical model that addresses key limitations of
the BTL model. Specifically, the statistical model accounts for judge
biases, captures the hierarchical structure of stimuli, incorporates
measurement error in trait estimation and hypothesis testing, and
accommodates violations of the equal dispersion assumption. These
improvements potentially enhance measurement reliability and validity
\citep{Perron_et_al_2015} and improve the statistical accuracy of
hypothesis testing.

@@

One promising research avenue involves investigating how the
hierarchical structure of stimuli affects CJ experiment designs and
hypotheses testing. One such study could examine whether collecting a
single text per individual provides sufficient information to produce
reliable trait estimates at different levels of the hierarchy. Due to
concerns about the feasibility of the comparison task
\citep{Boonen_et_al_2020}, many CJ studies assess traits using only one
text per individual \citep{citation}. This experimental design
implicitly assumes that within-individual variability is small relative
to between-individual differences--that is, texts produced by the same
individual are relatively homogeneous. However, researchers rarely test
this assumption, even though it directly influences decisions about the
number of texts and participants needed for the experiment, ultimately
affecting the reliability of the trait estimates.

Relatedly, another study could investigate hypotheses about the
characteristics of the texts themselves. Previously, the literature
lacked a direct and integrated method for evaluating these types of
hypotheses. Now, by collecting data on specific stimulus features,
researchers can assess whether these variables produce the expected
effects.

\section{Conclusion}\label{sec-conclusion}

>>>>

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key CJ experimental design features,
such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model to analyze dichotomous
pairwise comparison data. Finally, Section~\ref{sec-discussion}
discusses the findings, explores avenues for future research, and detail
the challenges for future researchers.

@@

Thurstone noted from the outset that his general form, described in
Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
problem}. Specifically, the model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep{Thurstone_1927b}. For instance, in a CJ experiment with five
texts, the general form would require estimating \(20\) parameters: five
modal discriminal processes, five discriminal dispersions, and \(10\)
correlations --one per comparison (see Table~\ref{tbl-thurstone_cases}).
However, a single judge could only provide \({5 \choose 2} = 10\) unique
comparisons, an insufficient data set to estimate the required
parameters.

To address this issue and facilitate the practical implementation of the
theory, Thurstone developed five cases derived from this general form,
each progressively incorporating additional simplifying assumptions
\citep{Thurstone_1927b}. In Case I, Thurstone postulated that pairs of
stimuli would maintain a constant correlation across all comparisons. In
Case II, he allowed multiple judges to undertake comparisons instead of
confining evaluations to a single judge. In Case III, he posited that
there was no correlation between stimuli. In Case IV, he assumed that
the stimuli exhibited similar dispersions. Finally, in Case V, he
replaced this assumption with the condition that stimuli had equal
discriminal dispersions. Table~\ref{tbl-thurstone_cases} summarizes the
assumptions of the general form and the five cases. For a detailed
discussion of these cases and their progression, refer to
\citet{Thurstone_1927b} and \citet[pp.~248--253]{Bramley_2008}.

@@

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement and offering no guidance on how to use
its trait estimates for statistical inference or hypothesis testing.
Specifically, Thurstone cautioned that its use ``should not be made
without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, as it
imposes the most extensive set of simplifying assumptions
\citep{Bramley_2008, Kelly_et_al_2022} (see
Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
goal was to produce a ``rather coarse scaling'' of traits and ``allocate
the compared stimuli on this continuum''
\citep[pp.~269]{Thurstone_1927b}, his theory did not support formal
statistical inference. Despite these limitations, it is surprising that
CJ research has predominantly relied on Case V to measure different
traits, which raises significant concerns about the reliability and
validity of such measurements in contexts where the case's assumptions
may not hold \citep{Kelly_et_al_2022, Andrich_1978}. Furthermore,
although the CJ tradition has attempted to address the gap of hypothesis
testing by using the point estimates of the traits--or their
transformations--a critical question remains: Does this approach provide
a valid foundation for statistical inference?

@@

As previously discussed, Case V remains the most widely used model in CJ
literature. This preference largely stems from the widespread adoption
of the BTL model, which provides a simplified statistical representation
of the case. The BTL model mirrors most of Case V's assumptions, with
one notable distinction: while Case V assumes a Normal distribution for
the stimuli' discriminal processes, the BTL model uses the more
mathematically tractable Logistic distribution
\citep{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
impact on trait estimation or model interpretation because the scale of
the discriminal process (i.e., the latent trait) is arbitrary up to a
non-monotonic transformation
\citep{vanderLinden_et_al_2017_I, McElreath_2021}. That is, as long as
the substitution (transformation) preserves the data rank order, the
choice of distribution for the discriminal processes is inconsequential.
This condition is satisfied in this case, as the Normal and Logistic
distributions exhibit analogous statistical properties, differing only
by a scaling factor of approximately \(1.7\)
\citep{vanderLinden_et_al_2017_I}.

@@

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent texts with different
dispersions or outlying observations, the common CJ practice of
excluding stimuli based on these statistics, as seen in
\citet{Pollitt_2012a}, \citet{Pollitt_2012b},
\citet{vanDaal_et_al_2016}, and \citet{Goossens_et_al_2018}, may
unintentionally discard valuable information, introducing bias into the
trait estimates \citep{Zimmerman_1994, McElreath_2020}. The direction
and magnitude of these biases remain unpredictable, as they depend on
which stimuli researchers exclude from the analysis.

@@

Building on the previous section, it is clear that, researchers in CJ
studies typically use the BTL model to measure traits and position the
compared stimuli along a latent continuum \citep{Thurstone_1927b}. The
CJ literature also show that they frequently rely on point estimates of
these traits (i.e, the BTL scores)--or their transformations--to conduct
statistical inference or hypothesis testing. For example, researchers
have used these scores to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct statistical inference or hypothesis tests, as this
practice can undermine the resulting statistical conclusions. A key
consideration is that BTL scores are parameter estimates that inherently
carry uncertainty (measurement error). Ignoring this uncertainty can
bias the analysis and reduce the precision of hypothesis tests. The
direction and magnitude of such biases are often unpredictable. Results
may be attenuated, exaggerated, or remain unaffected depending on the
degree of uncertainty in the scores and the actual effects being tested
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. Furthermore,
the reduced precision in hypothesis tests diminishes their statistical
power, increasing the likelihood of committing type-I or type-II errors
\citep{McElreath_2020}.

@@

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, the SCM incorporates a key design feature of CJ
experiments: the judges' biases \((B_{kj})\). This extension builds on
the arguments presented in Section~\ref{sec-theory-issue1b}, contending
that the discriminal difference becomes an observable outcome only
through judges' perceptions. Given that such perceptions may be
imperfect--and that each judge may carry some degree of bias
\citep[see][]{Pollitt_et_al_2003, vanDaal_et_al_2016}--it is reasonable
that judges' perceptions (bias) should be treated as an integral
component of the CJ system from the outset, as this leads to a more
accurate representation of the data-generating process underlying the
pairwise comparisons. This model defines the preliminary set of
endogenous variables,
\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\), and the
preliminary set of structural equations, \(F = { f_{O}, f_{D} }\), which
capture the non-parametric dependencies among these variables.

@@

Additionally, we construct the matrix \(R\) to map each row of the
\(IA\) matrix with a corresponding row from the \(JK\) matrix. This
matrix has \(n\) rows and \(6\) columns, where
\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
coefficient, which quantifies the total number of unique comparisons
possible between every pair of stimuli generated by each pair of
individuals in the population. Thus, we define the matrix as follows:

@@

In summary, the SCM~\ref{fig-cj15_scm} and DAG~\ref{fig-cj15_dag} extend
Thurstone's general form to address several limitations of the BTL
model. These extensions account for judge biases (see
Section~\ref{sec-theory-theoretical_P1}), reflect the hierarchical
structure of stimuli and incorporate measurement error in trait
estimation and hypothesis testing (see
Section~\ref{sec-theory-theoretical_P3}), and clarify the role of the
sample and comparison mechanisms in CJ experiments (see
Section~\ref{sec-theory-theoretical_SC}). However, they do not resolve
concerns about the assumption of equal dispersions among stimuli
discussed in Section~\ref{sec-theory-issue1a}. Since this concern
relates to the statistical assumption underlying the distribution of the
discriminal process, we develop a formal statistical model to address it
in the next section.

@@

Note that while the joint distribution in Equation (\ref{eq-mat51})
includes the probability distributions of the sampling and comparison
mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
\(P(Z_{J})\)--we omit these probabilities from the models presented in
Figure~\ref{fig-cj16}. We make this omission because, although these
distributions contribute to the overall joint distribution of the data,
they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
This independence stems from the random selection designs that govern
\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
\(Z_{J}\)\footnote{Randomization ensures that data---and, by extension,
  an estimator---satisfies several key identification properties, such
  as common support, no interference, and consistency. The most critical
  property, however, is the elimination of confounding.
  \emph{Confounding} occurs when an external variable, such as
  \(X_{I}\), simultaneously influences both the outcome (e.g.,
  \(O_{R}\)) and a variable of interest (e.g., \(S\)), resulting in
  spurious associations between the latter two
  \citep{Everitt_et_al_2010}. Randomization ensure the absence of
  confounding by effectively decoupling the association between the
  variable of interest and any other variable, except for the outcome
  itself. For a more detailed discussion on the benefits of
  randomization, see \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
  \citet{Neal_2020}, and \citet{Hernan_et_al_2020}.}.

Second, we define \(D_{R}\) as the difference between the discriminal
processes of two texts, \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\),
representing the underlying written-quality trait of the texts, plus the
corresponding repeated judge biases, \(B_{JK}[j, k]\). Note that if we
assume that \(B_{JK}[j,k]\) reflects the difference in stimulus-specific
biases, i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we
can re-write the discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation} This formulation reveals that the discriminal difference
captures a \emph{pure interaction effect}, in which neither the texts'
discriminal processes nor the judges' biases alone determine the
outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
terms, the discriminal processes of the stimuli become an observable
outcome only through the judges' perceptions (biases). To clarify, the
square brackets in \(D_{R}\) simply indicate the relevant indices for
each trait vector and do not imply any subsetting of the data.

@@

Fourth, we define \(T_{I}\) as a linear combination of the effects of
relevant student-related variables, \(\beta_{XI} X_{I}\), and students'
idiosyncratic errors, \(e_{I}\). As before, we assume that each element
of \(e_{I}\) is \(iid\) Normal with a mean of zero. However, we consider
the standard deviation, \(s_{XI}\), varies depending on the teaching
method group to which each student belongs. Given the example in
Section~\ref{sec-theoretical} where the teaching method
\(X_{I} = \{1,2\}\), the model sets the following constraint to anchor
the scale of the latent trait \citep{Depaoli_2021}:
\(\sum_{g=1}^{2} s_{XI}[g]/2 = 1\). Note that this parametrization of
\(e_{I}\) relaxes the equal dispersion assumption for the stimuli,
effectively addressing the concerns raised in
Section~\ref{sec-theory-issue1a}.

@@

Finally, we use \emph{Bayesian inference methods} to convert
model~\ref{fig-cj16_stat} into a practical statistical tool for
analyzing paired comparison data. Bayesian inference offers three key
advantages for this purpose. First, it handles complex and
overparameterized models, where the number of parameters exceeds the
number of observations \citep{Baker_1998, Kim_et_al_1999}--such as the
model we try to implement. Second, it incorporates prior information to
constrain parameter estimates within plausible bounds, thereby
mitigating estimation issues like non-convergence or improper solutions
that often affect frequentist methods
\citep{Martin_et_al_1975, Seaman_et_al_2011}. In our implementation, we
use priors to define the error distribution, which sets the scale of the
latent variables \citep{Depaoli_2014}. Third, Bayesian inference
supports robust inferences from small samples, where the asymptotic
properties underlying frequentist methods are less reliable
\citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}. This
feature is particularly relevant in CJ experiments, where researchers
often collect large volumes of paired comparisons but work with
relatively small samples of judges, stimuli, and individuals to test
hypotheses.

@@

Thurstone introduced the Law of Comparative Judgment to measure
psychological traits of stimuli through pairwise comparisons
\citep{Thurstone_1927a, Thurstone_1927b}. In its general form, the
theory models single-judge comparisons across multiple, potentially
correlated stimuli. Each comparison produces a dichotomous outcome
indicating which stimulus the judge perceives as having a higher trait
level. However, Thurstone identified one key issue in this model: it
required estimating more ``unknown'' parameters than the number of
available pairwise comparisons \citep{Thurstone_1927b}. To address this
issue and facilitate the theory's practical applicability, he formulated
five cases derived from the general form, each progressively
incorporating several simplifying assumptions.

Of these five cases, the CJ literature has predominantly relied on Case
V to measure various psychological traits, as shown in studies like
\citet{Kimbell_2012}, \citet{Jones_et_al_2015}, and
\citet{Boonen_et_al_2020}. This preference largely stems from the
widespread adoption of the BTL model, which offers a simplified
statistical formulation of the case. The BTL model mirrors the
assumptions of Case V--namely, equal discriminal dispersions and zero
correlation for the stimuli' discriminal processes--with one notable
distinction: while Case V assumes normally distributed discriminal
processes, the BTL model uses the more mathematically tractable Logistic
distribution \citep{Andrich_1978, Bramley_2008}.

Although replacing the Normal with the Logistic distribution has a
minimal impact on trait estimation or model interpretation
\citep{vanderLinden_et_al_2017_I, McElreath_2021}, Thurstone
acknowledged that the two remaining assumptions of Case V--and, by
extension, of the BTL model--could pose significant problems when
researchers assess complex traits or heterogeneous stimuli
\citep{Thurstone_1927b}. In particular, he noted that although these
assumptions simplify the trait measurement model, they could fail to
capture the complexity of CJ data
\citep{Kelly_et_al_2022, Andrich_1978}, potentially leading to
unreliable and inaccurate trait estimates
\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.

Furthermore, because Thurstone's primary goal was to produce a ``coarse
scaling'' of traits and to ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
how to use the resulting trait estimates for statistical inference or
hypothesis testing. The CJ tradition has attempted to address this gap
by separating trait estimation from hypothesis testing. Specifically, CJ
studies often rely on point estimates of the traits (i.e., the BTL
scores)--or their transformations--to conduct hypothesis tests. However,
the statistical literature cautions against this practice, as it can
introduce bias into the analysis and compromise the reliability of the
resulting conclusions
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

To address these issues, this study extends Thurstone's general form
using a systematic, integrated approach that combines Causal and
Bayesian inference methods. It begins by formulating a conceptual model
represented by an SCM and a DAG. This model integrates Thurstone's core
theoretical principles--such as the discriminal processes (i.e., traits)
of stimuli--alongside key CJ experimental design features, including
judges' bias, sampling procedures, and comparison mechanisms. The study
then translates the SCM into a bespoke statistical model that addresses
key limitations of the BTL model. In particular, it enables the analysis
of CJ data when the assumptions of equal dispersion and zero correlation
do not hold (see Section~\ref{sec-theory-issue1a} and
Section~\ref{sec-theory-issue1b}), and when researchers seek to conduct
statistical inference or hypothesis testing (refer to
Section~\ref{sec-theory-issue2}). These methodological innovations
potentially enhance the reliability and validity of trait measurement
\citep{Perron_et_al_2015} and improve the accuracy of statistical
conclusions.

@@

For instance,

Despite its potential, this study presents several challenges that
researchers must navigate. First, our approach does not establish a
clear epistemological connection between the theoretical latent trait
and its quantitative estimate. In other words, the approach does not
provide a definitive answer to the fundamental question: How do we know
that the estimated trait reflects the theoretical trait of interest?
However, this limitation is not unique to our approach; it reflects a
fundamental challenge all research involving latent constructs faces.
Nevertheless, we argue that our approach offers methodological tools
that enable researchers to design CJ experiments under conditions that
support reliable and valid trait measurement.

Second, although initial simulation tests of the statistical model have
yielded promising results (not shown), these findings may not generalize
across research contexts. We have argued that the model's ability to
estimate traits accurately and support valid statistical inferences
depends heavily on the specific research context--particularly on
whether the data and inferential goals align with the model's underlying
assumptions. That said, our approach provides tools for researchers to
test and evaluate these context-dependent CJ models directly.
